A novel physical mechanism for the spontaneous formation of porous crystals via viscoelastic phase separation

### 2016-07-19 crystal-gel/nature/crystal_gel3.tex

\title{A novel route to the spontaneous formation of porous crystals via viscoelastic phase separation}

@@

\date{Received November 26, 2015}

\begin{abstract}
{\bf
Viscoelastic phase separation of colloidal suspensions can be interrupted to form gels either by 
%the glass transition, with the formation of colloidal gels~\cite{poon2002,zaccarelli2007,piazza1994phase,verhaegh1997transient,tanaka1999colloid,lu2008gelation}, or by crystallization, which results in the formation of a crystalline network~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways,sabin2012}. 
glass transition~\cite{poon2002,zaccarelli2007,piazza1994phase,verhaegh1997transient,tanaka1999colloid,lu2008gelation} or by crystallization~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways,sabin2012}. 
%But while the former process has been studied in detail, the latter has never been observed experimentally at a single-particle level and it is thus lacking a microscopic understanding. 
With a new confocal microscopy protocol, we unveil a novel crystallization pathway to sponge-like porous structures in a colloid-polymer mixture undergoing viscoelastic gas-liquid phase separation. We show that crystallization leads to a network structure different from the original phase separation pattern. Crystals nucleate inside the liquid network, but grow past it by direct condensation of the gas phase on their surface, i.e. de-sublimation, driving evaporation of nearby liquid. %The latter processes via the gas phase has been overlooked so far. %This process represents the colloidal analogue of the Bergeron process~\cite{glickman2000glossary}, which explains the formation of ice crystals in mixed phase clouds and is at the origin of rain. Our finding gives us full experimental access to the kinetic pathway of this poorly understood phenomenon important for climate science. 
%The key is complex dynamic interplay involving the three phases upon simultaneous occurrence of two non-equilibrium phenomena, phase separation and crystallization. 
We argue that similar crystal-gel states can be formed as the result of the gas-liquid phase separation of any crystallizable components, such as monoatomic and molecular systems, if dynamics of the liquid phase is slow enough to induce viscoelastic phase separation, but fast enough to prevent immediate vitrification. This mechanism will provide a novel single-step pathway to form nano-porous crystals of metals and semiconductors without dealloying \cite{erlebacher2001evolution}, which may be important for catalytic, optical, sensing, and filtration applications \cite{ding2004nanoporous,ding2009nanoporous, wittstock2010nanoporous}.
%, but also sheds new light on crystal network formation in more complex systems such as magma \cite{philpotts1998role}, biominerals \cite{rousseau2005multiscale}, and foods \cite{deman1987fat}.}
}
\end{abstract}

@@

Unlike the above standard scenario of colloidal gelation, it is known that phase separation can also be arrested by crystallization, if the process takes place below the melting point of a component of a mixture~\cite{tanaka1985new}. Indeed the possibility of a different class of gels, which are stabilized by crystallization, 
was suggested by numerical simulations~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways} and observed in microgravity experiments \cite{sabin2012}. 
For colloidal systems, we can in principle access structural evolution in real time at a single-particle level; however, there has so far been no confocal microscopy  
study on the dynamical process of crystal-gel formation. 
For confocal microscopy experiments on colloid gelation, colloid-polymer mixtures have so far been used as a model system. 
The main difficulty consists in the preparation protocol of such a colloid-polymer mixture, where before observation of its dynamics shaking or applying shear is necessary to prepare an initial homogeneous state. This introduces turbulent flows at the beginning of the process, and also does not allow the
observation of the initial stages of crystallization. This initial perturbation may even alter the selection of the final non-equilibrium arrested state.  
In our experiments we instead succeed in building a novel protocol that allows the time evolution of the system to be observed directly
under the confocal microscope from the very early stages and in a quiescent situation without introducing fluid flows.

In this Letter, we show that, for low polymer concentration, the system can undergo a novel crystallization pathway that leads to the formation
of crystalline droplets. These droplets originate from crystal seeds that form inside the liquid branches of the spinodal aggregate of network morphology, and then grow by direct condensation from the gas phase. The process is analogous to the Bergeron process of ice formation in mixed phase clouds, where ice droplets grow at the
expense of the supercooled liquid droplets due to their lower saturated vapour pressure. 
This is speculated to be the primary mechanism for the formation of raindrops. 
Differently from the standard crystal-gel scenario, where crystals only nucleate and grow inside the dense branches of the gel, becoming dynamically arrested, we show that the kinetic
pathways involving the gas phase (predominantly the Bergeron process, but also Ostwald ripening) play a fundamental role. We also observe that increasing the strength of the interparticle attraction
(by increasing the polymer concentration), instead of forming crystal gels, the system undergoes the classical gelation process, in which
phase separation is interrupted by vitrification~\cite{verhaegh1997transient,tanaka1999colloid,Royall2007,lu2008gelation}.

@@

As described above and also in Methods, salt injection initiates liquid-gas phase separation of the colloidal suspension. 
All samples share the same early stages of spinodal decomposition. 
A typical early stage phase ordering process observed at $\phi\approx 0.33$ and $c_p=0.38$ mg/g can be seen in Supplementary Movie 1.. 
Due to strong dynamical asymmetry between colloids and the solvent \cite{tanaka1999colloid}, the colloidal particles start aggregating, 
eventually forming a percolating network. Thus, a dense network (liquid) coexists with freely diffusing monomers (gas).
In Fig.~\ref{fig:phase_separation}a, the state points which did not phase separate are indicated with open triangles, while
the other symbols indicate samples undergoing phase separation. \textcolor{blue}{The black dashed line is the gas spinodal computed from free volume theory and the 
black continuous line is the binodal line. They are in good agreement with our experimental points (the small discrepancy at small colloidal volume fractions 
is common in free volume theory)~\cite{Royall2007,lu2008gelation}.}
The topology of the network is similar to all our samples at the initial stages. 
This is shown for example in Fig.~\ref{fig:phase_separation}b, where we plot the radius of gyration of colloidal clusters ($R_g$)
as a function of the cluster size ($s$) at the percolation point. Two particles belong to the same cluster if their distance is
within the first peak of the radial distribution function. All samples show the same scaling law, $R_{\rm g}\sim s^{1/D}$, which is compatible with
the random gelation universality class exponent ($D=2.53$) as shown by the dashed line in Fig.~\ref{fig:phase_separation}b
(we do not imply that the formation of the network follows the random gelation universality class, just that our
fractal (or effective) dimension is compatible with it).

@@

\textcolor{blue}{To gain additional insight, in Supplementary Figure~\ref{S-fig:local_phi12} we plot the late stage (after all samples already underwent gas-liquid phase separation) time evolution of the local volume fraction of closed packed particles (having 12 neighbours). The volume fraction is obtained by computing the Voronoi diagram of each configuration, which uniquely assigns a volume to each colloidal particle. Only the state point at $\phi\approx 0.33$ and $c_p = 0.38\,$mg/g shows an increase, up to an asymptotic value that we identify with the composition of the stable crystal phase $\approx 0.72$. We plot the asymptotic average volume fraction as a function of the number of neighbours in Fig.~\ref{fig:phase_separation}d. Approaching close packing (12 neighbours), we clearly see two families of curves. The state point at $\phi\approx 0.33$ and $c_p=0.38\,$mg/g reaches close packing at the volume fraction $\approx 0.72$. By contrast all other state points, reach close packing at a markedly 
lower volume fraction, which is indeed close to the volume fraction of the attractive glass state~\cite{pham2002multiple}. This result clearly indicates that a single state point follows a different arrest mechanism, in which phase separation is arrested by crystallization and not by glassiness.}

@@

Supplementary Movie 2 shows the late stage (after the initial spinodal decomposition) evolution of a 2D confocal slice. Even on this raw data the growth of crystalline regions is obvious. More quantitatively, we adopt bond orientational analysis to detect which particles are in crystalline environments~\cite{russo2013interplay} 
\textcolor{blue}{(see Supplementary Information on the detail)}. 
With this analysis we reveal that crystallization events indeed start occurring for the state point $\phi\approx 0.33$ and $c_p=0.38$ mg/g, which has the lowest
polymer concentration of all state points. This is shown in Fig.~\ref{fig:crystals}a, where the black line indicates the average size of the crystallites, 
while the red line indicates the average number of crystallites as a function of time. The number of crystallites first rapidly increases as nucleation events
start occurring inside the liquid network, but eventually starts decreasing as the different crystallites grow and merge with each other.  \textcolor{blue}{Supplementary Movie 3 shows a three-dimensional (3D) reconstruction  of the whole phase ordering sequence that emphasizes crystalline particles.}
%The average size of the crystals shows an interesting linear growth regime for $100<t<600$. This linear growth regime goes past the time
%of coalescence of nuclei, so a first speculation is that it could be due to the compensation of the following two effects:
%\begin{itemize}
% \item diffusive growth of the nuclei, which scales like $t^{1/2}$
% \item coalescence growth, which should scale faster than $t$, maybe $t^2?$
%\end{itemize}
All other state points show only negligible signs of crystallization. In particular, for state points with $\phi\approx 0.33$, increasing
polymer concentration drastically reduces the amount of crystals. This is in agreement with the
idea of enhanced crystallization rates near metastable critical points~\cite{ten1997enhancement,olmsted1998spinodal} and the
results of Refs.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways},
which speculated two different arrest mechanisms: crystallization at low polymer concentration, and dynamic arrest at high polymer concentrations.

@@

\textcolor{blue}{Next we analyse in detail the crystallization trajectory, distinguishing between the different processes responsible for the
crystallization of the phase separated network.
%Supplementary Movies 2 and 3D show the time-lapse videos of 2D confocal microscopy observation and 3D computer reconstruction, respectively, for the crystallizing sample at $\phi\approx 0.33$ and $c_p=0.38$ mg/g,
%where the process of crystal growth by direct attachment of gas particles can be observed directly. 
From the trajectory, we extract slabs of 10 particle diameter thickness for configurations at different times, shown in Fig.~\ref{fig:transitions}a. 
Particles are coloured according to their phases (gas, liquid, and crystal). For the classification criteria, see Supplementary Information. The figure shows the first nucleation 
events inside the liquid network (left panel). At the same time liquid regions that have not crystallized start evaporating, and the gas phase contributes to the growth of nuclei (middle panel). Finally the different nuclei coalesce (right panel). 
Note that the topology of the crystal network found at late times, i.e. $t=5577\,\tau_{\rm B}$ (middle panel) and $t=12098\,\tau_{\rm B}$ (right panel) in Fig.~\ref{fig:transitions}a, differs significantly from that of the liquid network at early stages, i.e. $t=2055\,\tau_{\rm B}$ (left panel) in Fig.~\ref{fig:transitions}a. This suggests that the crystal network does not form only by direct freezing of the liquid, but also by other mechanisms which involve the evaporation of the liquid (Bergeron process) and the sublimation of small crystals (Ostwald ripening). In the following we will investigate these different crystallization mechanisms in depth.}

\textcolor{blue}{
In Fig.~\ref{fig:transitions}b, we show  the fraction of particles in each different phase for the crystallization trajectory at
$\phi\approx 0.33$ and $c_p=0.38$ mg/g, after the liquid-gas phase separation has occurred.
The process of crystallization is characterized 
by a steep decrease in liquid particles, as they transform into small crystalline nuclei inside the liquid domains. This decrease
is then accompanied by an increase in the fraction of gas particles: as the first crystals start to reach the gas phase, liquid particles evaporate 
to the gas phase due to the higher vapour pressure of the liquid phase compared to the crystalline phase. After the onset 
of a steady-state gas population, there are three growth channels for the crystal. Direct crystallization is the process by which 
crystals grow by incorporating nearby liquid particles. In the Bergeron process, liquid droplets first evaporate and the resulting gas phase 
contributes to the growth of crystalline regions. Ostwald ripening is instead the process by which small crystallites sublimate, and colloidal
particles are transferred to larger nuclei. In order to assess the relative importance of these three growth channels, we analyse the trajectory, and, for every time frame, assign each particle a state between gas, liquid and crystal. In order to minimize short-term fluctuations, the state of each colloidal particle is time averaged for $50\,\tau_{\rm B}$. We then measure the fraction of trajectories with different transition histories.
Direct crystallization accounts for 54\% of particles trajectories in which
gas or liquid particles transition to the crystal state without liquid evaporation or crystal de-sublimation. The Bergeron process accounts for 35\% of particle trajectories in which liquid particles
transition to the gas state before crystallizing. Ostwald ripening accounts for 11\% of particle trajectories in which crystal particles transition to the gas state before returning
to the crystal state. The different crystallization channels are depicted in the diagram of Fig.~\ref{fig:transitions}c.}

\textcolor{blue}{
In Supplementary Information we confirm these results with a detailed analysis of transition probabilities, including a state for surface particles that are in contact 
to crystalline particles, see Supplementary Figure~\ref{S-fig:probabilities}.
While the direct freezing of the fluid represents the major contribution to the nuclei growth, the kinetic path via the gas phase (gas$\rightarrow$crystal), also plays a crucial role,
especially in determining the morphology of the porous crystal, as we discussed above. In this context the Bergeron process~\cite{glickman2000glossary,morrison2012resilience}
plays a considerably more important role than Ostwald ripening.}

@@

The process of formation of ``crystal gels'' may be generic to many other 
systems. The requirements are (i) the presence of gas-liquid phase separation below the melting point of a crystal, (ii) 
weak or little frustration against crystallization (in our case, the use of monodisperse colloids), 
(iii) dynamical slowing down in a supercooled liquid state, which is necessary to induce viscoelastic phase separation leading to the formation 
of a network structure of the minority liquid phase, and (iv) the degree of supercooling is low enough to avoid a high rate of densification of the liquid phase 
leading to vitrification.  
Many monoatomic and single-component molecular systems can satisfy all these conditions in a certain range of the temperature and pressure. 
This can be seen, for example, by looking at the phase diagram of a Lennard-Jones liquid \cite{lodge1997brownian}, which represents many molecular systems 
without specific directional interactions. \textcolor{blue}{For monoatomic systems such as noble metals, condition (iii) may not be satisfied easily. 
To access glassy slow dynamics in a liquid phase, we may need a deep quench at a high pressure. However, we note that even without strong dynamic asymmetry 
due to glassiness, bicontinuous phase separation can take place between 35-85 volume \% of the liquid phase in ordinary gas/liquid phase separation \cite{Onuki2002}; 
and, thus, porous crystalline structures can be formed, although a thin network structure may be difficult to be formed.}

@@

\textcolor{blue}{Matching phase boundaries often leads to better determination of volume fractions than colloid diameter measurements~\cite{Poon2012}. We obtain the best match between theoretical phase diagram and experimental data for $\sigma=2.25\,\mu$m and $R_g=80$ nm (see Supplementary method). Therefore the polymer-colloid size ratio is $q_R=2R_g/\sigma=0.071$ and the overlap mass fraction of polymer $2.25$~mg/g.}

@@

\textcolor{blue}{
The crystal-gel state is reproduced in a sealed cell with all the components of the suspension pre-mixed. The colloid volume fraction is virtually identical at $\approx 0.33$ and polymer concentration is $0.40\,$mg/g. See Supplementary Figure~\ref{S-fig:capillary}.}

@@

\begin{figure*}[!t]
 \centering
 \includegraphics{phase_separation.pdf}
 \caption{{\bf Different routes of viscoelastic phase separation.} 
{\bf a,} Phase diagram. Empty black triangles are experimental point showing a fluid behaviour. The other symbols are the experimental points showing gel behaviour, consistently used in all panels. \textcolor{blue}{The dotted lines are the gas-liquid tie lines determined experimentally, with arrow heads indicating measured gas and liquid compositions (see Supplementary Information on the procedure to extract the compositions from confocal images).  Gray areas are theoretical equilibrium fluid and crystal regions. The solid and dashed lines are the metastable gas-liquid binodal and spinodal respectively.}
{\bf b,} Radius of gyration ($R_g$) as a function of cluster size ($s$) for all state points at percolation. The
 gray line represents the fractal dimension of the random gelation universality class.
{\bf c,} Structure factors for the late stages of the gelation process. The gray and black lines represents respectively random gelation and compact fractal dimensions.
\textcolor{blue}{{\bf d,} Local volume fraction around a colloidal particle, as a function of the number of nearest neighbours of the particle for the state points in the very late stage. 
The two horizontal straight lines indicate the characteristic volume fractions in attractive systems, crystal ($\sim 0.72$) and glass ($\sim 0.64$). }
 }
 \label{fig:phase_separation}
\end{figure*}

@@

\begin{figure*}[!t]
 \centering
 \includegraphics[width=12cm]{fig2}
 % crystal_size.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=(atend)
\caption{\textcolor{blue}{{\bf Percolated network structures.} {\bf a:} gel; {\bf b:} crystal-gel. The gel state is obtained from the state
point $\phi\approx 0.33$ and $c_p=1.07$ mg/g at $t=400$ min after sample preparation, while the crystal-gel is obtained at the same $\phi$ but
at lower polymer concentration, $c_p=0.38$ mg/g at $t=468$ min after sample preparation. On the left, the network is represented as a continuous field that has
been obtained by applying a Gaussian filter to the position of the particles (of width equal to the inter-particle distance) and
plotting the surface which bounds the highest 20 \% values of the field. We can clearly see not only the percolated nature of both networks, 
but also the difference in the smoothness of the network surface between amorphous and crystal gels. 
On the right, slabs of 4.5 $\mu$m thickness and 41 $\mu$m width are plotted,
with particles coloured with the same colour scheme as in Fig.~\ref{fig:transitions}: red (crystal), green (gas particles), white (liquid particles). 
We note that very few liquid particles remain and the network is almost crystalline. 
}} 
\label{fig:network}
\end{figure*}

>>>>

\title{A novel physical mechanism for the spontaneous formation of porous crystals via viscoelastic phase separation}

@@

\date{Received July 20, 2016}

\begin{abstract}
{\bf
Viscoelastic phase separation of colloidal suspensions can be interrupted to form gels either by 
%the glass transition, with the formation of colloidal gels~\cite{poon2002,zaccarelli2007,piazza1994phase,verhaegh1997transient,tanaka1999colloid,lu2008gelation}, or by crystallization, which results in the formation of a crystalline network~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways,sabin2012}. 
glass transition~\cite{poon2002,zaccarelli2007,piazza1994phase,verhaegh1997transient,tanaka1999colloid,lu2008gelation} or by crystallization~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways,sabin2012,zhang2012non}. 
%But while the former process has been studied in detail, the latter has never been observed experimentally at a single-particle level and it is thus lacking a microscopic understanding. 
With a new confocal microscopy protocol, we unveil a novel crystallization pathway to sponge-like porous crystal structures in a colloid-polymer mixture undergoing viscoelastic gas-liquid phase separation. We show that this crystallization pathway leads to a network structure different from the original phase separation pattern. Crystals nucleate inside the liquid network, but grow past it by direct condensation of the gas phase on their surface, i.e. de-sublimation, driving evaporation of nearby liquid. %The latter processes via the gas phase has been overlooked so far. %This process represents the colloidal analogue of the Bergeron process~\cite{glickman2000glossary}, which explains the formation of ice crystals in mixed phase clouds and is at the origin of rain. Our finding gives us full experimental access to the kinetic pathway of this poorly understood phenomenon important for climate science. 
%The key is complex dynamic interplay involving the three phases upon simultaneous occurrence of two non-equilibrium phenomena, phase separation and crystallization. 
We argue that similar crystal-gel states can be formed as the result of the gas-liquid phase separation of any crystallizable components, such as monoatomic and molecular systems, if dynamics of the liquid phase is slow enough to induce viscoelastic phase separation, but fast enough to prevent immediate vitrification. This mechanism will provide a novel single-step pathway to form nano-porous crystals of metals and semiconductors without dealloying \cite{erlebacher2001evolution}, which may be important for catalytic, optical, sensing, and filtration applications \cite{ding2004nanoporous,ding2009nanoporous, wittstock2010nanoporous}.
%, but also sheds new light on crystal network formation in more complex systems such as magma \cite{philpotts1998role}, biominerals \cite{rousseau2005multiscale}, and foods \cite{deman1987fat}.}
}
\end{abstract}

@@

Unlike the above standard scenario of colloidal gelation, it is known that phase separation can also be arrested by crystallization, if the process takes place below the melting point of a component of a mixture~\cite{tanaka1985new}. Indeed the possibility of a different class of gels, which are stabilized by crystallization, 
was suggested by numerical simulations~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways} and observed in experiments \cite{sabin2012,zhang2012non}. 
For colloidal systems, we can in principle access structural evolution in real time at a single-particle level; however, there has so far been no confocal microscopy  
studies on the dynamical process of crystal-gel formation on a microscopic level, which makes the microscopic mechanism of crystal-gel formation elusive. 
For confocal microscopy experiments on colloid gelation, colloid-polymer mixtures have so far been used as a model system. 
The main difficulty consists in the preparation protocol of such a colloid-polymer mixture, where before observation of its dynamics shaking or applying shear is necessary to prepare an initial homogeneous state. This introduces turbulent flows at the beginning of the process, and also does not allow the
observation of the initial stages of crystallization. This initial perturbation may even alter the selection of the final non-equilibrium arrested state.  
In our experiments we instead succeed in building a novel protocol that allows the time evolution of the system to be observed directly
under the confocal microscope from the very early stages and in a quiescent situation without introducing fluid flows.

In this Letter, we show that, for low polymer concentration, the system can undergo a novel crystallization pathway that leads to the formation
of crystalline droplets. These droplets originate from crystal seeds that form inside the liquid branches of the spinodal aggregate of network morphology, and then grow by direct condensation from the gas phase. The process is analogous to the Bergeron process of ice formation in mixed phase clouds, where ice droplets grow at the
expense of the supercooled liquid droplets due to their lower saturated vapour pressure. 
This is speculated to be the primary mechanism for the formation of raindrops. 
Differently from the standard crystal-gel scenario, where crystals only nucleate and grow inside the dense branches of the gel, becoming dynamically arrested, we show that the kinetic
pathways involving the gas phase (predominantly the Bergeron process, but also Ostwald ripening) play a fundamental role. 
We reveal that the final crystal-gel network structure has smoother interface than the ordinary gels formed by vitrification because of this novel crystal growth mechanism.  
We also observe that increasing the strength of the interparticle attraction
(by increasing the polymer concentration), instead of forming crystal gels, the system undergoes the classical gelation process, in which
phase separation is interrupted by vitrification~\cite{verhaegh1997transient,tanaka1999colloid,Royall2007,lu2008gelation}.

@@

As described above and also in Methods, salt injection initiates liquid-gas phase separation of the colloidal suspension. 
All samples share the same early stages of spinodal decomposition. 
A typical early stage phase ordering process observed at $\phi\approx 0.33$ and $c_p=0.38$ mg/g can be seen in Supplementary Movie 1.  
Due to strong dynamical asymmetry between colloids and the solvent \cite{tanaka1999colloid}, the colloidal particles start aggregating, 
eventually forming a percolating network. Thus, a dense network (liquid) coexists with freely diffusing monomers (gas).
In Fig.~\ref{fig:phase_separation}a, the state points which did not phase separate are indicated with open triangles, while
the other symbols indicate samples undergoing phase separation. The black dashed line is the gas spinodal computed from free volume theory and the 
black continuous line is the binodal line. They are in good agreement with our experimental points (the small discrepancy at small colloidal volume fractions 
is common in free volume theory)~\cite{Royall2007,lu2008gelation}.
The topology of the network is similar to all our samples at the initial stages. 
This is shown for example in Fig.~\ref{fig:phase_separation}b, where we plot the radius of gyration of colloidal clusters ($R_g$)
as a function of the cluster size ($s$) at the percolation point. Two particles belong to the same cluster if their distance is
within the first peak of the radial distribution function. All samples show the same scaling law, $R_{\rm g}\sim s^{1/D}$, which is compatible with
the random gelation universality class exponent ($D=2.53$) as shown by the dashed line in Fig.~\ref{fig:phase_separation}b
(we do not imply that the formation of the network follows the random gelation universality class, just that our
fractal (or effective) dimension is compatible with it).

@@

To gain additional insight, in Supplementary Figure~\ref{S-fig:local_phi12} we plot the late stage (after all samples already underwent gas-liquid phase separation) time evolution of the local volume fraction of closed packed particles (having 12 neighbours). The volume fraction is obtained by computing the Voronoi diagram of each configuration, which uniquely assigns a volume to each colloidal particle. Only the state point at $\phi\approx 0.33$ and $c_p = 0.38\,$mg/g shows an increase, up to an asymptotic value that we identify with the composition of the stable crystal phase $\approx 0.72$. We plot the asymptotic average volume fraction as a function of the number of neighbours in Fig.~\ref{fig:phase_separation}d. Approaching close packing (12 neighbours), we clearly see two families of curves. The state point at $\phi\approx 0.33$ and $c_p=0.38\,$mg/g reaches close packing at the volume fraction $\approx 0.72$. By contrast all other state points, reach close packing at a markedly 
lower volume fraction, which is indeed close to the volume fraction of the attractive glass state~\cite{pham2002multiple}. This result clearly indicates that a single state point follows a different arrest mechanism, in which phase separation is arrested by crystallization and not by glassiness.

@@

Supplementary Movie 2 shows the late stage (after the initial spinodal decomposition) evolution of a 2D confocal slice. Even on this raw data the growth of crystalline regions is obvious. More quantitatively, we adopt bond orientational analysis to detect which particles are in crystalline environments~\cite{russo2013interplay} 
(see Supplementary Information on the detail). 
With this analysis we reveal that crystallization events indeed start occurring for the state point $\phi\approx 0.33$ and $c_p=0.38$ mg/g, which has the lowest
polymer concentration of all state points. This is shown in Fig.~\ref{fig:crystals}a, where the black line indicates the average size of the crystallites, 
while the red line indicates the average number of crystallites as a function of time. The number of crystallites first rapidly increases as nucleation events
start occurring inside the liquid network, but eventually starts decreasing as the different crystallites grow and merge with each other.  Supplementary Movie 3 shows a three-dimensional (3D) reconstruction  of the whole phase ordering sequence that emphasizes crystalline particles. Although crystal clusters are isolated in this movie, they are bridged by the dense colloidal liquid phase and involved in the percolated gel network, as can be seen in Fig. \ref{fig:network}b. 
We can see that crystal clusters grow by condensation from the colloidal gas phase. 
%The average size of the crystals shows an interesting linear growth regime for $100<t<600$. This linear growth regime goes past the time
%of coalescence of nuclei, so a first speculation is that it could be due to the compensation of the following two effects:
%\begin{itemize}
% \item diffusive growth of the nuclei, which scales like $t^{1/2}$
% \item coalescence growth, which should scale faster than $t$, maybe $t^2?$
%\end{itemize}
All other state points show only negligible signs of crystallization. In particular, for state points with $\phi\approx 0.33$, increasing
polymer concentration drastically reduces the amount of crystals. This is in agreement with the
idea of enhanced crystallization rates near metastable critical points~\cite{ten1997enhancement,olmsted1998spinodal} and the
results of Refs.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways},
which speculated two different arrest mechanisms: crystallization at low polymer concentration, and dynamic arrest at high polymer concentrations.

@@

Next we analyse in detail the crystallization trajectory, distinguishing between the different processes responsible for the
crystallization of the phase separated network.
%Supplementary Movies 2 and 3D show the time-lapse videos of 2D confocal microscopy observation and 3D computer reconstruction, respectively, for the crystallizing sample at $\phi\approx 0.33$ and $c_p=0.38$ mg/g,
%where the process of crystal growth by direct attachment of gas particles can be observed directly. 
From the trajectory, we extract slabs of 10 particle diameter thickness for configurations at different times, shown in Fig.~\ref{fig:transitions}a. 
Particles are coloured according to their phases (gas, liquid, and crystal). For the classification criteria, see Supplementary Information. The figure shows the first nucleation 
events inside the liquid network (left panel). At the same time liquid regions that have not crystallized start evaporating, and the gas phase contributes to the growth of nuclei (middle panel). Finally the different nuclei coalesce (right panel). 
Note that the topology of the crystal network found at late times, i.e. $t=5577\,\tau_{\rm B}$ (middle panel) and $t=12098\,\tau_{\rm B}$ (right panel) in Fig.~\ref{fig:transitions}a, differs significantly from that of the liquid network at early stages, i.e. $t=2055\,\tau_{\rm B}$ (left panel) in Fig.~\ref{fig:transitions}a. This suggests that the crystal network does not form only by direct freezing of the liquid, but also by other mechanisms which involve the evaporation of the liquid (Bergeron process) and the sublimation of small crystals (Ostwald ripening). In the following we will investigate these different crystallization mechanisms in depth.

In Fig.~\ref{fig:transitions}b, we show  the fraction of particles in each different phase for the crystallization trajectory at
$\phi\approx 0.33$ and $c_p=0.38$ mg/g, after the liquid-gas phase separation has occurred.
The process of crystallization is characterized 
by a steep decrease in liquid particles, as they transform into small crystalline nuclei inside the liquid domains. This decrease
is then accompanied by an increase in the fraction of gas particles: as the first crystals start to reach the gas phase, liquid particles evaporate 
to the gas phase due to the higher vapour pressure of the liquid phase compared to the crystalline phase. After the onset 
of a steady-state gas population, there are three growth channels for the crystal. Direct crystallization is the process by which 
crystals grow by incorporating nearby liquid particles. In the Bergeron process, liquid droplets first evaporate and the resulting gas phase 
contributes to the growth of crystalline regions. Ostwald ripening is instead the process by which small crystallites sublimate, and colloidal
particles are transferred to larger nuclei. In order to assess the relative importance of these three growth channels, we analyse the trajectory, and, for every time frame, assign each particle a state between gas, liquid and crystal. In order to minimize short-term fluctuations, the state of each colloidal particle is time averaged for $50\,\tau_{\rm B}$. We then measure the fraction of trajectories with different transition histories.
Direct crystallization accounts for 54\% of particles trajectories in which
gas or liquid particles transition to the crystal state without liquid evaporation or crystal de-sublimation. The Bergeron process accounts for 35\% of particle trajectories in which liquid particles
transition to the gas state before crystallizing. Ostwald ripening accounts for 11\% of particle trajectories in which crystal particles transition to the gas state before returning
to the crystal state. The different crystallization channels are depicted in the diagram of Fig.~\ref{fig:transitions}c.

In Supplementary Information we confirm these results with a detailed analysis of transition probabilities, including a state for surface particles that are in contact 
to crystalline particles, see Supplementary Figure~\ref{S-fig:probabilities}.
While the direct freezing of the fluid represents the major contribution to the nuclei growth, the kinetic path via the gas phase (gas$\rightarrow$crystal), also plays a crucial role,
especially in determining the morphology of the porous crystal, as we discussed above. In this context the Bergeron process~\cite{glickman2000glossary,morrison2012resilience}
plays a considerably more important role than Ostwald ripening.

@@

The process of formation of ``crystal gels'' may be generic to many other 
systems. The requirements are (i) the presence of gas-liquid phase separation below the melting point of a crystal, (ii) 
weak or little frustration against crystallization (in our case, the use of monodisperse colloids), 
(iii) dynamical slowing down in a supercooled liquid state, which is necessary to induce viscoelastic phase separation leading to the formation 
of a network structure of the minority liquid phase, and (iv) the degree of supercooling is low enough to avoid a high rate of densification of the liquid phase 
leading to vitrification.  
Many monoatomic and single-component molecular systems can satisfy all these conditions in a certain range of the temperature and pressure. 
This can be seen, for example, by looking at the phase diagram of a Lennard-Jones liquid \cite{lodge1997brownian}, which represents many molecular systems 
without specific directional interactions. For monoatomic systems such as noble metals, condition (iii) may not be satisfied easily. 
To access glassy slow dynamics in a liquid phase, we may need a deep quench at a high pressure. However, we note that even without strong dynamic asymmetry 
due to glassiness, bicontinuous phase separation can take place between 35-85 volume \% of the liquid phase in ordinary gas/liquid phase separation \cite{Onuki2002}; 
and, thus, porous crystalline structures can be formed, although a thin network structure may be difficult to be formed.

@@

Matching phase boundaries often leads to better determination of volume fractions than colloid diameter measurements~\cite{Poon2012}. We obtain the best match between theoretical phase diagram and experimental data for $\sigma=2.25\,\mu$m and $R_g=80$ nm (see Supplementary method). Therefore the polymer-colloid size ratio is $q_R=2R_g/\sigma=0.071$ and the overlap mass fraction of polymer $2.25$~mg/g.

@@

The crystal-gel state is reproduced in a sealed cell with all the components of the suspension pre-mixed. The colloid volume fraction is virtually identical at $\approx 0.33$ and polymer concentration is $0.40\,$mg/g. See Supplementary Figure~\ref{S-fig:capillary}.

@@

\begin{figure*}[!t]
 \centering
 \includegraphics{phase_separation.pdf}
 \caption{{\bf Different routes of viscoelastic phase separation.} 
{\bf a,} Phase diagram. Empty black triangles are experimental point showing a fluid behaviour. The other symbols are the experimental points showing gel behaviour, consistently used in all panels. The dotted lines are the gas-liquid tie lines determined experimentally, with arrow heads indicating measured gas and liquid compositions (see Supplementary Information on the procedure to extract the compositions from confocal images).  Gray areas are theoretical equilibrium fluid and crystal regions. The solid and dashed lines are the metastable gas-liquid binodal and spinodal respectively.
{\bf b,} Radius of gyration ($R_g$) as a function of cluster size ($s$) for all state points at percolation. The
 gray line represents the fractal dimension of the random gelation universality class.
{\bf c,} Structure factors for the late stages of the gelation process. The gray and black lines represents respectively random gelation and compact fractal dimensions.
{\bf d,} Local volume fraction around a colloidal particle, as a function of the number of nearest neighbours of the particle for the state points in the very late stage. 
The two horizontal straight lines indicate the characteristic volume fractions in attractive systems, crystal ($\sim 0.72$) and glass ($\sim 0.64$). 
 }
 \label{fig:phase_separation}
\end{figure*}

@@

\begin{figure*}[!t]
 \centering
 \includegraphics[width=12cm]{fig2}
 % crystal_size.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=(atend)
\caption{{\bf Percolated network structures.} {\bf a:} gel; {\bf b:} crystal-gel. The gel state is obtained from the state
point $\phi\approx 0.33$ and $c_p=1.07$ mg/g at $t=400$ min after sample preparation, while the crystal-gel is obtained at the same $\phi$ but
at lower polymer concentration, $c_p=0.38$ mg/g at $t=468$ min after sample preparation. On the left, the network is represented as a continuous field that has
been obtained by applying a Gaussian filter to the position of the particles (of width equal to the inter-particle distance) and
plotting the surface which bounds the highest 20 \% values of the field. We can clearly see not only the percolated nature of both networks, 
but also the difference in the smoothness of the network surface between amorphous and crystal gels. 
On the right, slabs of 4.5 $\mu$m thickness and 41 $\mu$m width are plotted,
with particles coloured with the same colour scheme as in Fig.~\ref{fig:transitions}: red (crystal), green (gas particles), white (liquid particles). 
We note that very few liquid particles remain and the network is almost crystalline. 
} 
\label{fig:network}
\end{figure*}
